Words aren't created equal: Investigating bias on the CDI

### 2021-10-29 paper/cogsci.Rmd
wrote an abstract

abstract: >
    Include no author information in the initial submission, to facilitate
    blind review.  The abstract should be one paragraph, indented 1/8 inch on both sides,
    in 9~point font with single spacing. The heading 'Abstract'
    should be 10~point, bold, centered, with one line of space below
    it. This one-paragraph abstract section is required only for standard
    six page proceedings papers. Following the abstract should be a blank
    line, followed by the header 'Keywords' and a list of
    descriptive keywords separated by semicolons, all in 9~point font, as
    shown below.

keywords: >
    language development;

@@

- introduce importance of measuring early language development (e.g., links to later educational outcomes)
- introduce CDI
- predictive validity: Fenson et al. 1994, Bornstein et al. 2012, Duff et al. 2015

- discuss potential for bias
- what does between-group bias look like?

- IRT - what is a good measure?

@@

- what is our goal for measuring vocabulary? - identifying language delays, predicting later reading, or talking...
- how you select items influences how well you achieve these goals (and what bias you find)
We take the approach put forward by @stenhaug2021treading: GLIMMER plots based on the 1-parameter logistic are sufficient to identify bias, without potentially obscuring DIF in a more complex model's additional parameters.

>>>>

abstract: >
    Children's early language skill has been linked to later educational outcomes, making
    it important to accurately measure early language. Parent-reported instruments such as the 
    Communicative Development Inventories (CDIs) have been shown to provide valid, consistent 
    measures of children's aggregate early language skill. However, CDIs are predominantly 
    comprised of hundreds of vocabulary items, some of which may not be heard (and thus learned) 
    equally often by children of varying backgrounds. Here, we use a database of American English 
    CDIs to identify words that show strong bias for particular groups of children, along the axes
    of sex (male vs. female), race/ethnicity (white vs. non-white), and socioeconomic status 
    (high vs. low). For each axis, we identify dozens of strongly biased items, and show that 
    eliminating these items reduces the expected ability difference between groups. For sex, we
    consider how to propose replacement words that may show less bias, on the basis of their 
    relatively equal frequency in adult speech directed to male and female children.

keywords: >
    language development; early vocabulary; measuring instrument bias;

@@

introduce importance of measuring early language development (e.g., links to later educational outcomes)

introduce CDI

predictive validity: Fenson et al. 1994, Bornstein et al. 2012, Duff et al. 2015

discuss potential for bias

what does between-group bias look like?

IRT - what is a good measure?

@@

what is our goal for measuring vocabulary? - identifying language delays, predicting later reading, or talking...

how you select items influences how well you achieve these goals (and what bias you find)

We take the approach put forward by @stenhaug2021treading: GLIMMER plots based on the 1-parameter logistic are sufficient to identify bias, without potentially obscuring DIF in a more complex model's additional parameters.

### 2021-10-29 paper/cogsci.tex
wrote an abstract

\title{Investigating Item Bias}

@@

\begin{abstract}
Include no author information in the initial submission, to facilitate
blind review. The abstract should be one paragraph, indented 1/8 inch on
both sides, in 9\textasciitilde point font with single spacing. The
heading `Abstract' should be 10\textasciitilde point, bold, centered,
with one line of space below it. This one-paragraph abstract section is
required only for standard six page proceedings papers. Following the
abstract should be a blank line, followed by the header `Keywords' and a
list of descriptive keywords separated by semicolons, all in
9\textasciitilde point font, as shown below.

\textbf{Keywords:}
language development;
\end{abstract}

@@

\begin{itemize}
\item
  introduce importance of measuring early language development (e.g.,
  links to later educational outcomes)
\item
  introduce CDI
\item
  predictive validity: Fenson et al.~1994, Bornstein et al.~2012, Duff
  et al.~2015
\item
  discuss potential for bias
\item
  what does between-group bias look like?
\item
  IRT - what is a good measure?
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  what is our goal for measuring vocabulary? - identifying language
  delays, predicting later reading, or talking\ldots{}
\item
  how you select items influences how well you achieve these goals (and
  what bias you find) We take the approach put forward by
  (\textbf{Stenhaug2021?}): GLIMMER plots based on the 1-parameter
  logistic are sufficient to identify bias, without potentially
  obscuring DIF in a more complex model's additional parameters.
\end{itemize}

@@

We analyze American English CDI: Words \& Sentences administrations from
5520 children aged 16 to 30 from Wordbank (\textbf{Frank2016?};
\textbf{Frank2021?}).

@@

Research was conducted over 5520 participants, whose data was parent
reported and collected by the Wordbank Project. Unlike other projects
Wordback relies on the kindness of others to contribute their data which
means that often meta-data for some sets is missing. Even so, the
project now has data over 29 different languages on children from ages
8-30 months old, though our study focused entirely on American English
and Children 16-30 months. When they received complete information, the
Wordbank data set collected demographic data consisting of Birth Order,
Race/Ethnicity, Sex and Mothers Education. For our study we focused on
the comparing data between demographic groups on the axis of Sex,
Ethnicity and Mothers Education, which was a proxy for social economic
status. Out data included 1989 female and 2105 male participants, and an
additional 1426 participants whose sex remained unreported. In terms of
Ethnicity data was recorded from 2202 white, 67 Asian, 222 Black, 131
Hispanic, and 93 ``Other'' participants as well as 2805 unreported.
Given our distribution of ethnicities, for our analysis we split
participants into White (2202) and Non-White (513). Finally, we split
participants into high and low SES, of which we had NA high SES
participants, corresponding to mothers' education of some college or
higher and NA Low SES participants.

>>>>

\title{Words aren't created equal: Investigating bias on the CDI}

@@

\begin{abstract}
Children's early language skill has been linked to later educational
outcomes, making it important to accurately measure early language.
Parent-reported instruments such as the Communicative Development
Inventories (CDIs) have been shown to provide valid, consistent measures
of children's aggregate early language skill. However, CDIs are
predominantly comprised of hundreds of vocabulary items, some of which
may not be heard (and thus learned) equally often by children of varying
backgrounds. Here, we use a database of American English CDIs to
identify words that show strong bias for particular groups of children,
along the axes of sex (male vs.~female), race/ethnicity (white
vs.~non-white), and socioeconomic status (high vs.~low). For each axis,
we identify dozens of strongly biased items, and show that eliminating
these items reduces the expected ability difference between groups. For
sex, we consider how to propose replacement words that may show less
bias, on the basis of their relatively equal frequency in adult speech
directed to male and female children.

\textbf{Keywords:}
language development; early vocabulary; measuring instrument bias;
\end{abstract}

@@

introduce importance of measuring early language development (e.g.,
links to later educational outcomes)

introduce CDI

predictive validity: Fenson et al.~1994, Bornstein et al.~2012, Duff et
al.~2015

discuss potential for bias

what does between-group bias look like?

IRT - what is a good measure?

@@

what is our goal for measuring vocabulary? - identifying language
delays, predicting later reading, or talking\ldots{}

how you select items influences how well you achieve these goals (and
what bias you find)

We take the approach put forward by Stenhaug, Frank, \& Domingue (2021):
GLIMMER plots based on the 1-parameter logistic are sufficient to
identify bias, without potentially obscuring DIF in a more complex
model's additional parameters.

Differential item functioning (DIF) is a technique in IRT used to
identify items that show bias against demographic groups. DIF is a
statistical characteristic of an item that shows the extent to which the
item might operate differently or measure varying abilities for
subgroups and members of separate demographic groups. DIF is, however, a
challenging metric for identifying bias for a couple reasons. Firstly,
DIF is extremely hard to accurately root out. The process of finding DIF
involves using the test you used to do IRT analysis in order to look for
bias within the very same test. Secondly, the presence of DIF does not
necessarily indicate the presence of bias; it can indicate that one
demographic has a higher average ability level than the other. These two
issues are the basis upon which we get the Fundamental DIF
Identification problem.

This problem can be be understood through a concrete example that is
well-known from research on early language learning: consider the fact
that females show a larger vocabulary than males across early
development {[}see Frank, Braginsky, Yurovsky, \& Marchman (2021) Ch.
6?; OTHER REFS{]}. The question however is whether or not girls actually
have a higher ability level or if the tests have words that are bias
toward girls ie. are easier for girls, or perhaps the test is biased
toward boys but the natural ability of girls is enough to overcome that
bias. To answer this question we need to know the ability levels of
girls and boys in order to confirm that if a word is learned earlier by
girls it is simply because of an ability difference rather than a bias
inherent to the word. The problem arises from the fact that we measure
ability level using the very same test that we are trying to check for
biased words. If the boys and girls were of the same ability level this
wouldn't be an issue but given the evidence showing that girls have a
higher ability level, we need to know the difference in ability so that
when we find DIF for a specific word can be sure it is outside of the
expecting DIF that results naturally from their difference in ability.

@@

We analyze American English CDI: Words \& Sentences administrations from
5520 children aged 16 to 30 from Wordbank (Frank, Braginsky, Yurovsky,
\& Marchman, 2017; Frank et al., 2021).

@@

Research was conducted over 5520 participants, whose data was parent
reported and collected by the Wordbank Project. Unlike other projects
Wordbank relies on the kindness of others to contribute their data which
means that often meta-data for some sets is missing. Even so, the
project now has data over 29 different languages on children from ages
8-30 months old, though our study focused entirely on American English
and Children 16-30 months. When they received complete information, the
Wordbank data set collected demographic data consisting of Birth Order,
Race/Ethnicity, Sex and Mothers Education. We focused on comparing data
between demographic groups on the axes of Sex, Ethnicity and Mother's
Education, a proxy for socioeconomic status (henceforth, SES). Our data
included 1989 female and 2105 male participants, and an additional 1426
participants whose sex remained unreported. In terms of Ethnicity data
was recorded from 2202 white, 67 Asian, 222 Black, 131 Hispanic, and 93
``Other'' participants as well as 2805 unreported. Given our
distribution of ethnicities, for our analysis we split participants into
White (2202) and Non-White (513). Finally, we split participants into
high and low SES, of which we had 4973 high SES participants,
corresponding to mothers' education of some college or higher and 547
Low SES participants.

\hypertarget{one-column-images}{%
\subsection{One-column images}\label{one-column-images}}

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

Place acknowledgments (including funding information) in a section at
the end of the paper.

\hypertarget{references}{%
\section{References}\label{references}}

\setlength{\parindent}{-0.1in} 
\setlength{\leftskip}{0.125in}

\noindent

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Frank2017}{}%
Frank, M. C., Braginsky, M., Yurovsky, D., \& Marchman, V. A. (2017).
Wordbank: An open repository for developmental vocabulary data.
\emph{Journal of Child Language}, \emph{44}(3), 677--694.

\leavevmode\hypertarget{ref-Frank2021}{}%
Frank, M. C., Braginsky, M., Yurovsky, D., \& Marchman, V. A. (2021).
\emph{Variability and consistency in early language learning: The
wordbank project}. MIT Press.

\leavevmode\hypertarget{ref-stenhaug2021treading}{}%
Stenhaug, B., Frank, M. C., \& Domingue, B. (2021). Treading carefully:
Agnostic identification as the first step of detecting differential item
functioning.
